A standard for the scholarly citation of archaeological data as an incentive to data sharing

### 2017-12-28 analysis/paper/paper.Rmd
small requests from editor, reference list

```{r sharing-rates}
tbl <- readr::read_csv("../figures/tbl_studies_of_data_sharing.csv")
knitr::kable(tbl, caption = "Summary of a selection of previous studies of data sharing in various fields, from Spencer (2010). In the study by Peng et al., the authors did not attempt to contact other researchers, but noted the availability of data as described in the published article.")
```

```{r date-and-sharing, fig.cap="There is no obvious trend in data sharing becavious and date of publication"}
library(viridis)
library(scales) 
library(wesanderson)

@@

library(cowplot)
email_sharing_plots <-
plot_grid(sharing_by_year,
                     sharing_by_student,
                     sharing_by_country,
                     sharing_by_journal,
          axis = "lt",
          ncol = 2,
          align = "hv",
          labels = "AUTO")

@@

```{r email-figs, fig.cap="Summary of results of email requests for data. A. Responses by year of journal article publication, B. Student status of first author, C. Country location of first author, D. Journal (AAS = Archaeological and Anthropological Sciences, AQ = American Antiquity, ARCHMTRY = Archaeometry, JAS = Journal of Archaeological Science, JFA = Journal of Field Archaeology, JHE = Journal of Human Evolution, LT = Lithic Technology, SCI = Science."}
knitr::include_graphics("../figures/email_sharing_plots_hand_edit.png")
```

@@

@@

In addition to our email survey, we also conducted a pilot study of a random sample of `r nrow(jas_article_survey)` 48 articles published during February-May 2017 in the _Journal of Archaeological Science_ to investigate data sharing behaviors. In this sample we found openly available raw data in `r data_are_avail` papers (`r data_are_avail_perc`%), even though only `r have_data_avail_statement` articles (`r have_data_avail_statement_perc`%) include a data availability statement. For papers where data was available, `r data_in_supp` papers (`r data_in_supp_perc`%) have data contained in supplementary files published with the paper, `r data_in_table` (`r data_in_table_perc`%) have all the raw data in tables in the text of the paper, and `r data_in_repo` (`r data_in_repo_perc`%) have data in an online repository. The most frequently shared type of data is compositional (e.g. element concentrations, n = `r comp_n`, `r comp_perc`%), followed by radiocarbon and luminescence age data (where the dates are the primary object of analysis, n = `r dates_n`, `r dates_perc`%) and DNA sequences (n = `r dna_n`, `r dna_perc`%). Generally, compositional data is presented in tables in the text, dating data is presented in supplementary files, and DNA data are in repositories (as is typical for DNA data) (Figure \@ref(fig:data-types-and-files)). Data shared in supplementary files and online repositories were most often in Microsoft Excel format (n = `r xlsx_n`, `r xlsx_perc`% of the files in supplements files and repositories), followed by CSV (comma separated variables, an open source plain text spreadsheet format, n = `r csv_n`, `r csv_perc`%), and tables embedded in Microsoft Word or PDF documents (n = `r doc_n`, `r doc_perc`%). When a scripting language such as R, Python, MATLAB or OxCal was used as the analysis software for a paper (n = `r script_n`, `r script_perc`% of the papers where would identify the software), only 2 papers also provided the script with the paper.

@@

```{r data-types-and-files, fig.cap="Summary of relationships between file types, data types, and data locations. A. File types and data locations, B. Data types and data locations. "}
plot_data_type <- plot_grid(plot_file_type_location,
                            plot_data_type_location,
                      labels="AUTO", ncol = 1)
plot_data_type
```

@@

# change in each major repo over time
per_repo <-
dates  %>%
  filter(!is.na(yyyy_date)) %>%
  group_by(yyyy_date, datacentre)  %>%
  tally() %>%
  filter(n != 0,
         datacentre %in% top_data_sources[1:9]) %>%
    mutate(datacentre = str_wrap(datacentre, width = )) %>%
   ggplot(aes(yyyy_date,
              n
             )) +
  geom_line(size = 1)  +
  scale_fill_manual(values = wes_palette("Rushmore")) +
  scale_y_log10() +
  xlim(c(1990, 2017)) +
  xlab("Year DOI registered (1990-2017 only)") +
  ylab("Number of DOIs\nregistered per year") +
  theme_new() +
  facet_wrap(~ datacentre)

# show plots
datacite_temporal_plots <-
 plot_grid(date_plot,
           per_repo,
           labels="AUTO",
           ncol = 1,
           align = "hv",
           axis = "lr")

@@

```{r datacite-temporal-summary, fig.cap = "Change over time in archaeology datasets known to DataCite. A. Counts of datasets per year by the major data centres, red lines indicate major increases in the number of datasets in repositories. Note that the date on the horizontal axis indicates when the data was collected, not when it was deposited in the data centre, B. Temporal trends in the number of DOIs at individual data centres."}
knitr::include_graphics("../figures/datacite_temporal_plots.png")
```

@@

```{r, data-centre-file-types, fig.cap = "Relationships between data centres and file formats. The ten most frequently found file formats are shown here, in the data centres with these files from the top ten data centres by number of DOIs. Open file formats for structured data, such as CSV, are not common."}
knitr::include_graphics("../figures/plot_structured_data_files_per_file_tally.png")
```

@@

Alsheikh-Ali A.A., Qureshi W., Al-Mallah M.H. and J. Ioannidis
2011. Public Availability of Published Research Data in High-Impact Journals. PLOS ONE 6(9): e24357. DOI:10.1371/journal.pone.0024357, accessed 7 September 2017.
American Psychological Association
2001.	Ethical Standards for the Reporting and Publishing of Scientiﬁc Information. In Publication Manual edited by American Psychological Association, pp. 387–396. Washington, DC
ARCHES project
n.d. The Standard and Guide to Best Practice in Archaeological Archiving in Europe. Electronic document, http://archaeologydataservice.ac.uk/arches/Wiki.jsp?page=ARCHES%20Introduction, accessed on 7 September 2017.
Bogen J, and J. Woodward
1998.	Saving the Phenomena. The Philosophical Review 97(3): 303–352
Borgman, C. L. 
2015.	Big Data, Little Data, No Data: Scholarship in the Networked World. MIT Press, Massachusetts. 
Borgman, C. L. 
2012.	The Conundrum of Sharing Research Data. Journal of the American Society for Information Science and Technology, 63(6), 1059–1078.
Chambers, C. 
2017.	The Seven Deadly Sins of Psychology: A Manifesto for Reforming the Culture of Scientific Practice. Princeton University Press, Princeton, New Jersey. 
Clarke, M.
2015.	The Digital Dilemma: Preservation and the Digital. Archaeological Record. Advances in Archaeological Practice. 3(4): 313-330.
Cranston K, Harmon L. J., O'Leary, M. A. and C. Lisle
2014.	Best Practices for Data Sharing in Phylogenetic Research. PLOS Currents Tree of Life. (blog) 19 Jun 2014, DOI: 10.1371/currents.tol.bf01eff4a6b60ca4825c69293dc59645, accessed 7 September 2017
Crossref 
2017. Event Data User Guide, Version 1.0.0 - Beta. Electronic document, https://www.eventdata.crossref.org/guide/, accessed 7 November  2017.
Daston, L, and P. Galison
1992.	 The Image of Objectivity. Representations. 40:81–128.
Data Citation Synthesis Group
2014.	Joint Declaration of Data Citation Principles. In FORCE11; 2014 edited by Martone M., Electronic document, https://doi.org/10.25490/a97f-egyk, accessed on 7 September 2017. 
Dorch, B. F., Drachen, T. M., and O. Ellegaard
2015.	The Data Sharing Advantage in Astrophysics. Proceedings of the International Astronomical Union, 11(A29A), 172-175.

>>>>

```{r sharing-rates}
tbl <- readr::read_csv("../figures/tbl_studies_of_data_sharing.csv")
knitr::kable(tbl, caption = "Summary of a selection of previous studies of data sharing in various fields, from Spencer (2010). In the study by Peng et al. (2006), the authors did not attempt to contact other researchers, but noted the availability of data as described in the published article.")
```

```{r date-and-sharing, fig.cap="There is no obvious trend in data sharing behavious and date of publication"}
library(viridis)
library(scales) 
library(wesanderson)

@@

library(cowplot)
email_sharing_plots <-
plot_grid(sharing_by_year,
                     sharing_by_student,
                     sharing_by_country,
                     sharing_by_journal,
          axis = "lt",
          ncol = 2,
          align = "hv",
          labels = letters[1:4])

@@

```{r email-figs, fig.cap="Summary of results of email requests for data: a. Responses by year of journal article publication, b. Student status of first author, c. Country location of first author, d. Journal (AAS = Archaeological and Anthropological Sciences, AQ = American Antiquity, ARCHMTRY = Archaeometry, JAS = Journal of Archaeological Science, JFA = Journal of Field Archaeology, JHE = Journal of Human Evolution, LT = Lithic Technology, SCI = Science."}
knitr::include_graphics("../figures/email_sharing_plots_hand_edit.png")
```

@@

total_number_of_articles_published_in_the_population_that_we_drew_our_sample <- 48

@@

In addition to our email survey, we also conducted a pilot study of a random sample of `r total_number_of_articles_published_in_the_population_that_we_drew_our_sample` articles published during February-May 2017 in the _Journal of Archaeological Science_ to investigate data sharing behaviors. In this sample we found openly available raw data in `r data_are_avail` papers (`r data_are_avail_perc`%), even though only `r have_data_avail_statement` articles (`r have_data_avail_statement_perc`%) include a data availability statement. For papers where data was available, `r data_in_supp` papers (`r data_in_supp_perc`%) have data contained in supplementary files published with the paper, `r data_in_table` (`r data_in_table_perc`%) have all the raw data in tables in the text of the paper, and `r data_in_repo` (`r data_in_repo_perc`%) have data in an online repository. The most frequently shared type of data is compositional (e.g. element concentrations, n = `r comp_n`, `r comp_perc`%), followed by radiocarbon and luminescence age data (where the dates are the primary object of analysis, n = `r dates_n`, `r dates_perc`%) and DNA sequences (n = `r dna_n`, `r dna_perc`%). Generally, compositional data is presented in tables in the text, dating data is presented in supplementary files, and DNA data are in repositories (as is typical for DNA data) (Figure \@ref(fig:data-types-and-files)). Data shared in supplementary files and online repositories were most often in Microsoft Excel format (n = `r xlsx_n`, `r xlsx_perc`% of the files in supplements files and repositories), followed by CSV (comma separated variables, an open source plain text spreadsheet format, n = `r csv_n`, `r csv_perc`%), and tables embedded in Microsoft Word or PDF documents (n = `r doc_n`, `r doc_perc`%). When a scripting language such as R, Python, MATLAB or OxCal was used as the analysis software for a paper (n = `r script_n`, `r script_perc`% of the papers where would identify the software), only 2 papers also provided the script with the paper.

@@

```{r data-types-and-files, fig.cap="Summary of relationships between file types, data types, and data locations: a. File types and data locations, b. Data types and data locations. "}
plot_data_type <- plot_grid(plot_file_type_location,
                            plot_data_type_location,
                      labels=letters[1:2], ncol = 1)
plot_data_type
```

@@

# change in each major repo over time
labels <- c("Archaeology Data Service",
            "Digital Antiquity",
            "DANS-EASY",
            "TOPOI-ExcellenzCluster",
            "Swedish National Data Service",
            "figshare",
            "ZENODO",
            "ResearchGate",
            "Eurographics",
            "University Library Heidelberg")
names(labels) <- top_data_sources[1:9]

per_repo <-
dates  %>%
  filter(!is.na(yyyy_date)) %>%
  group_by(yyyy_date, datacentre)  %>%
  tally() %>%
  filter(n != 0,
         datacentre %in% top_data_sources[1:9]) %>%
    mutate(datacentre = str_wrap(datacentre, width = )) %>%
   ggplot(aes(yyyy_date,
              n
             )) +
  geom_line(size = 1)  +
  scale_fill_manual(values = wes_palette("Rushmore")) +
  scale_y_log10() +
  xlim(c(1990, 2017)) +
  xlab("Year DOI registered (1990-2017 only)") +
  ylab("Number of DOIs\nregistered per year") +
  theme_new() +
  facet_wrap(~ datacentre, labeller = labeller(datacentre = labels[1:9]))

# show plots
datacite_temporal_plots <-
 plot_grid(date_plot,
           per_repo,
           labels=letters[1:2],
           ncol = 1,
           align = "hv",
           axis = "lr")

@@

```{r datacite-temporal-summary, fig.cap = "Change over time in archaeology datasets known to DataCite: a. Counts of datasets per year by the major data centers, red lines indicate major increases in the number of datasets in repositories. Note that the date on the horizontal axis indicates when the data was collected, not when it was deposited in the data center, b. Temporal trends in the number of DOIs at individual data centers."}
knitr::include_graphics("../figures/datacite_temporal_plots.png")
```

@@

```{r, data-centre-file-types, fig.cap = "Relationships between data centers and file formats. The ten most frequently found file formats are shown here, in the data centers with these files from the top ten data centers by number of DOIs. Open file formats for structured data, such as CSV, are not common."}
knitr::include_graphics("../figures/plot_structured_data_files_per_file_tally.png")
```

@@

Alsheikh-Ali, Alawi A., Waqas Qureshi, Mouaz H. Al-Mallah and John P. A. Ioannidis
2011 Public Availability of Published Research Data in High-Impact Journals. _PLOS ONE_ 6(9): e24357. DOI:10.1371/journal.pone.0024357, accessed 7 September 2017.

American Psychological Association
2001	Ethical Standards for the Reporting and Publishing of Scientiﬁc Information. In _Publication Manual edited by American Psychological Association_, pp. 387–396. Washington, DC

ARCHES project
n.d. _The Standard and Guide to Best Practice in Archaeological Archiving in Europe._ Electronic document, http://archaeologydataservice.ac.uk/arches/Wiki.jsp?page=ARCHES%20Introduction, accessed on 7 September 2017.

Bogen, James, and James Woodward
1998	Saving the Phenomena. _The Philosophical Review_ 97(3): 303–352

Borgman, Christine L. 
2015	_Big Data, Little Data, No Data: Scholarship in the Networked World._ MIT Press, Massachusetts.

Borgman, Christine L. 
2012	The Conundrum of Sharing Research Data. _Journal of the American Society for Information Science and Technology_ 63(6): 1059–1078.

Chambers, Chris 
2017	_The Seven Deadly Sins of Psychology: A Manifesto for Reforming the Culture of Scientific Practice._ Princeton University Press, Princeton, New Jersey.

Clarke, Mary E.
2015	The Digital Dilemma: Preservation and the Digital. Archaeological Record. _Advances in Archaeological Practice_ 3(4): 313-330.

Cranston, Karen, Luke J. Harmon, Maureen A. O'Leary, and Curtis Lisle
2014	Best Practices for Data Sharing in Phylogenetic Research. _PLOS Currents Tree of Life._ (blog) 19 June 2014, DOI: 10.1371/currents.tol.bf01eff4a6b60ca4825c69293dc59645, accessed 7 September 2017

Crossref 
2017  Event Data User Guide, Version 1.0.0 - Beta. Electronic document, https://www.eventdata.crossref.org/guide/, accessed 7 November 2017.

Daston, LLorraine, and Peter Galison
1992	The Image of Objectivity. _Representations_ 40:81–128.

Data Citation Synthesis Group
2014	Joint Declaration of Data Citation Principles. In _FORCE11; 2014_ edited by M. Martone, Electronic document, https://doi.org/10.25490/a97f-egyk, accessed on 7 September 2017.

Dorch, B. F., Drachen, T. M., and O. Ellegaard
2015	The Data Sharing Advantage in Astrophysics. Proceedings of the International Astronomical Union, 11(A29A), 172-175.
